<!-- rmarkdown v1 -->

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)

---

[![minimal R version](https://img.shields.io/badge/R%3E%3D-3.6.0-6666ff.svg)](https://cran.r-project.org/)
[![packageversion](https://img.shields.io/badge/Package%20version-1.0.0-orange.svg?style=flat-square)](commits/master)

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
```{r, include=FALSE}
devtools::load_all(".")
```

# Overview of the fairness R package

The fairness R package provides a tool to easily calculate algorithmic fairness metrics for given predicted probabilities or predicted classes between sensitive groups of populations. It also provides additional opportunities to compare various other prediction metrics between subgroups. 

You can load the package by running:
```{r eval = FALSE}
install.packages("fairness")
library(fairness)
```

## The fairness pipeline

You will find a detailed tutorial in the missCompare vignette. We recommend that you spend some time going through the vignette, as it contains a much more in-depth description of the fairness package compared to this brief readme.

```{r eval = FALSE}
vignette("fairness")
```

## Description

This package contains functions to compute the most commonly used metrics of algorithmic fairness such as:   

- Demographic parity
- Proportional parity
- Equalized odds
- Predictive rate parity

In addition, the following comparisons are also implemented:    

- False positive rate parity
- False negative rate parity
- Accuracy parity
- Negative predictive value parity
- Specificity parity
- ROC AUC comparison
- MCC comparison

Most fairness measures are computed based on the confusion matrix resulting from a model fit of a given classification model.

## Brief tutorial

### Loading the COMPAS sample data set

```{r eval = TRUE}
data("compas")
```

The data already contains all variables necessary to run all parity metrics. In case you set up your own predictive model, you will need to concatenate predicted probabilities or predictions (0/1) to your original dataset.

### Running a selected function

```{r eval = FALSE}
prop_parity(data=compas, 
           group="ethnicity",
           probs="probability", 
           preds = NULL,
           outcome_levels = c("no","yes"), 
           cutoff = 0.5, 
           base = "Caucasian")
```

### Taking a look at the output

```{r eval = TRUE, echo = FALSE}
output <- prop_parity(data=compas, 
           group="ethnicity",
           probs="probability", 
           preds = NULL,
           outcome_levels = c("no","yes"), 
           cutoff = 0.5, 
           base = "Caucasian")
```

Metrics for propor parity:     
```{r , echo = FALSE, fig.width=5, fig.height=3}
output$Metric
```

Bar chart for the demographic parity metric:
```{r , echo = FALSE, fig.width=5, fig.height=3}
output$Metric_plot
```

Predicted probability plot for all subgroups:
```{r , echo = FALSE, fig.width=5, fig.height=3}
output$Probability_plot
```

## Acknowledgments
- Calders, T., & Verwer, S. (2010). Three naive Bayes approaches for discrimination-free classification. Data Mining and Knowledge Discovery, 21(2), 277-292.
- Chouldechova, A. (2017). Fair prediction with disparate impact: A study of bias in recidivism prediction instruments. Big data, 5(2), 153-163.
- Feldman, M., Friedler, S. A., Moeller, J., Scheidegger, C., & Venkatasubramanian, S. (2015, August). Certifying and removing disparate impact. In Proceedings of the 21th ACM SIGKDD International Conference on Knowledge Discovery and Data Mining (pp. 259-268). ACM.
- Friedler, S. A., Scheidegger, C., Venkatasubramanian, S., Choudhary, S., Hamilton, E. P., & Roth, D. (2018). A comparative study of fairness-enhancing interventions in machine learning. arXiv preprint arXiv:1802.04422.
- Zafar, M. B., Valera, I., Gomez Rodriguez, M., & Gummadi, K. P. (2017, April). Fairness beyond disparate treatment & disparate impact: Learning classification without disparate mistreatment. In Proceedings of the 26th International Conference on World Wide Web (pp. 1171-1180). International World Wide Web Conferences Steering Committee.


## Issues, questions

In case you need help or advice on fairness metrics or you want to report an issue, please do so in a reproducible example at the [fairness GitHub](https://github.com/kozodoi/Fairness/issues) page.

